---
title:  ""
author: ""
date: ""
header-includes:
   - \usepackage{animate}
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, eval=FALSE, warning = FALSE, message = FALSE, fig.align ="center")

```

#Data Visualizations - Base `R`

***

This module will cover: 

  - plotting in base R
  
which will require the following skills already covered:

  - reading in data
  - manipulating data  
  - assigning an object


##Introduction

Data visualization in `R` can be quick and dirty (ie. data exploration for yourself) or highly polished (ie. communicating to others). We have already touched on quick data exploration in the third module yesterday. Today we will cover content to produce a more polished looking plot. 

##Two approaches to the same problem: Base `R` vs `ggplot`

Plotting in base `R` can allow the user to create highly customized plots. This customization takes time, and requires many decisions. An alternative is to use the package `ggplot2` developed by Hadley Wickham based on the [*Grammar of Graphics*](http://www.springer.com/us/book/9780387245447) written by Leland Wilkinson. `ggplot2` has its own unique syntax that is a bit different from base `R`. We will walk through an example using base `R` and then recreate the figure using `ggplot2`. For even more side-by-side examples, see Nathan Yau's blog post on [Flowing Data](https://flowingdata.com/2016/03/22/comparing-ggplot2-and-r-base-graphics/). 

### Base `R`

Let's create a plot of the total population by county area for 5 midwest states (example taken from Selva Prabhakaran's [tutorial](http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html#Scatterplot_)). This data is part of the `ggplot2` package. We start with the basic scatterplot function `plot()` and then customize from there. 
```{r base R, eval=TRUE}
library(ggplot2) #load the package with the data
data("midwest", package = "ggplot2") #load the data, midwest is now in the working environment. 
plot(y=log10(midwest$poptotal), x=midwest$area) #call the x and y values 
     
```


A simple plot can take many more lines of code than you expect based on the visualization. When plotting in base `R` you'll use a handful of parameter settings in either `par()` or in the plotting related functions listed below. 

#### Important parameters ###
* **pch**: plotting symbol
* **lty**: line type
* **lwd**: line width
* **col**: plotting color
* **las**: orientation of axis labels 
* **bg**: background color
* **mar**: margin size
* **oma**: outer margin size
* **xlab**: X axis label
* **ylab**: Y axis label
* **main**: plot title
* **mfrow**: number of plots per row, column. Plots filled in row-wise.
* **mfcol**: number of plots per row, column. Plots filled in column-wise.



 
```{r base R customizations, eval=TRUE}
library(ggplot2) #load the package with the data
data("midwest", package = "ggplot2") #load the data, midwest is now in the working environment. 
plot(y=log10(midwest$poptotal), x=midwest$area, #call the x and y values 
     col=as.factor(midwest$state), #point colors should be based on state 
     pch=19, cex=.75,#point shape and size
     ylim=c(3,7), xlim=c(0,.1), #set the axis limites
     las=1, #rotate the axis labels 
     xlab="Area", ylab=expression('Log'[10]*'(Total population)'),#label the axis
     main ="Area vs population"#add a title
     )

```


####Legends and more

This is where the true power of plotting with base `R` customization shows. You can change the axis ticks and lables, add text anywhere, and even create multiple figures in a single visualization. The most common addition to any visualization will be the legend since they are not automatically created when plotting with base `R`. You have to add them manually. There are a few different methods to do this, but the function `legend()` works in most cases. To add the legend to the plot above, run the `legend()` function following the `plot()` function. 

```{r legend}
legend("topright", col=c(1:5), pch=19,legend=levels(as.factor(midwest$state)))
```

The visualization would then look like this: 

```{r legend added, eval=TRUE, echo=FALSE}
plot(y=log10(midwest$poptotal), x=midwest$area, #call the x and y values 
     col=as.factor(midwest$state), #point colors should be based on state 
     pch=19, cex=.75,#point shape and size
     ylim=c(3,7), xlim=c(0,.1), #set the axis limites
     las=1, #rotate the axis labels 
     xlab="Area", ylab=expression('Log'[10]*'(Total population)'),#label the axis
     main ="Area vs population"#add a title
     )
legend("topright", col=c(1:5), pch=19,legend=levels(as.factor(midwest$state)))
```


#### Other important functions ###
* **lines**: add lines to plot
* **points**: add points to plot
* **text**: add text
* **title**: add title to axes or plot
* **mtext**: add margin text
* **axis**: add axis tick/labels

####A quick note on multipanel plots

A grid of plots in base `R` can be created using parameter setting mfrow or cfrow. Base `R` also gives you the option to make inset or subplots like this example here where the boxplot is inside the histogram. 

```{r inset plot, eval=TRUE}
x <- rnorm(100,sd=0.5)  #generate random data to plot
y <- rbinom(100, 1, 0.5) #generate random data to plot
par(fig = c(0,1,0,1)) #set dimensions of histogram figure from bottom, left, top, and right

hist(x)  #plot main figure

par(fig = c(0.07,0.35, 0.5, 1), new = T) #set dimensions of inset plot  
boxplot(x ~ y) #plot inset

```

The `layout()` function allows the user to create multipanel plots of different sizes, like this: 

```{r layout plot, eval=TRUE}
# One figure in row 1 and two figures in row 2
# row 1 is 1/3 the height of row 2
# column 2 is 1/4 the width of the column 1 
attach(mtcars)
layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE), 
  	widths=c(3,1), heights=c(1,2))
hist(wt)
hist(mpg)
hist(disp)

```
####**Practice**: Make your own figure in Base `R`

Use the midwest data that you loaded above to make an original graphic. Use at least 2 of the following functions:

* **plot**: make a scatterplot
* **hist**: make a histogram
* **barplot**: make a barplot
* **countor**: make a contour plot
* **lines**: add lines to plot
* **points**: add points to plot
* **curve**: plot a function curve
* **text**: add text
* **title**: add title to axes or plot
* **mtext**: add margin text
* **axis**: add axis tick/labels

And at least 2 of the following parameters:

* **pch**: plotting symbol
* **lty**: line type
* **lwd**: line width
* **col**: plotting color
* **las**: orientation of axis labels 
* **bg**: background color
* **mar**: margin size
* **oma**: outer margin size
* **xlab**: X axis label
* **ylab**: Y axis label
* **main**: plot title
* **mfrow**: number of plots per row, column. Plots filled in row-wise.
* **mfcol**: number of plots per row, column. Plots filled in column-wise.

```{r Custom plot}
data("midwest", package = "ggplot2") #load the data, midwest is now in the working environment. 


```

Be prepared to share your plot and how you made it with a partner or with the class!


If you're interested in other customizations in base `R` check out Paul Murrell's [R Graphics](https://www.stat.auckland.ac.nz/~paul/RG2e/) book. 
&nbsp;
