---
title: "5-Data-Exploration"
author: "Kate Sabey"
date: "10/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Cleaning and Exploration

This module will cover:
* loading data
* calculating simple stats in base R
*	merging data sets

Along the way, we'll practice the following skills we already covered:
*	logic statements
*	manipulating data
*	assigning objects

Before analyzing your data, you'll need to start with:
1. Data cleaning (ie. removing missing values) should be done in R so you have a record of how the data was changed. 
2. Data exploration is a key step in scientific inquiry. Effective exploration might include transforming the data
   (e.g. by grouping replicate populations, calculating summary statistics, etc. 
3. Data visualization is also an important step to understanding your data, which we'll delve into more later (Section 6).

## Loading data

In addition to importing data from an external source (e.g. an xlsx or csv file), we can also load data from a package. Here, we'll explore the `flights` data in the data package called `nycflights13`, which contains airline on-time data for all flights leaving airports in the New York City, NY area in 2013.

*Let's install this package by using `install.packages()` command then load it using the `library()` command, as we learned yesterday.*

```{r load flights data}

install.packages("nycflights13") # install package
library(nycflights13) # load library
```

In addition to `flights`, you will notice the `nycflights13` package also includes other data sets including:
*	airlines
*	airports
*	planes
*	weather
You can access them specifically using double colons following the package name `nycflights13::`, but they are also loaded in the workspace automatically.

## Exploring Data Structure

In data exploration, we will first describe the broad trends in the data by understanding the structure of your data set.

*Write code below to answer the following questions:*
  *1. What is the data structure? How many columns and rows are there, and what data is represented in them?*
  *2. How many total flights and how many total flights per airline carrier?*
  *3. Create a new data set containing just United Airlines flights*
  *4. What are the different flight destinations?*

```{r exploring data structure}

# 1. What is the data structure? How many columns and rows are there, and what data is represented in them?
View(flights) # take a look at the whole flight data set as a separate tab in R studio
str(flights) # 336776 rows (each row is a flight) and 19 columns, confirm column classes (e.g. numeric versus factor)
names(flights) # returns names of columns
summary(flights) # returns relevant stats for each column

# 2. How many total flights and how many total flights per airline carrier?
carrier_table <- flights %>% 
  group_by(carrier) %>% 
  summarize(n()) # returns total flights (rows) per airline carrier
carrier_table

?table # this can also be accomplished using the table() function...
carrier_table <- table(flights$carrier)
carrier_table

carrier_table <- c() # ...or using a for loop!
for(i in unique(flights$carrier)){
  carrier_table[i] <- nrow(subset(flights, carrier == i))
}
carrier_table

# 3. Create a new data set containing just United Airlines flights
flights_UA <- flights %>%
  filter(carrier == "UA") # data set with only UA flights
flights_UA
nrow(flights_UA) # returns the number of United Airlines flights (~58K)

flights_UA <- subset(flights, carrier == "UA") # alternative option

flights_no_UA <- flights %>%
  filter(carrier != "UA") # data set with all flights except UA

# 4. What are the different flight destinations?
flights_dest <- unique(flights$dest) # create vector of airport destinations using the unique function
flights_dest # view airport destinations
length(flights_dest) # get number of airport destinations (105)
```


## Exploring Data with Summary Statistics

Next, we can try to answer more specific questions related to our potential research goals by using simple summary statistics.
Descriptive statistics built into R include: `mean`, `sd`, `var`, `min`, `max`, `median`, `range`, and `quantile`

*Write code below to answer the following questions:*
  *1. What is the max, min, and mean flight arrival delay? How does the mean arrival delay break down by airline?*
  *2. What is the most common flight destination?*
  *3. What is the number of flights per month?*

```{r exploring summary statistics}

# 1. What is the max, min, and mean flight arrival delay? How does the mean arrival delay break down by airline?

?max # check to determine function for excluding NAs in calculation
max(flights$arr_delay, na.rm = T)
min(flights$arr_delay, na.rm = T)
mean(flights$arr_delay, na.rm = T)

summary(flights$arr_delay) # if we don't need the values individually, can also extract them from the summary() output

flights %>%
  group_by(carrier) %>%
  summarize(mean_delay = mean(arr_delay, na.rm = T))

summary(UA$arr_delay) # delay statistics for a single airline (UA in this case)

# 2. What is the most common flight destination?
flights %>%
  group_by(dest) %>%
  summarize(number_flights = n()) %>%
  arrange(desc(number_flights)) # arranges in descending order - ORD = most common

# 3. What is the number of flights per month?
flights %>%
  group_by(month) %>%
  summarize(number_flights = n())

```



