---
title: "07-Data-Visualization-ggplot2"
author: "Kate Sabey"
date: "10/28/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A more detailed look at ggplot2

`ggplot2` is a great package for quickly making attractive and informative plots in R.
Builidng on the practice you had earlier, we'll first go step-by-step through formatting a nice scatterplot in `ggplot2`. 
Then we'll split up into small groups to explore making different types of plots using new ggplot "geoms".
Finally, I'll show off a couple of other useful visualization tools in `ggplot2` and adjacent packages!

*We can make the exact same scatterplot as we did with base graphics using ggplot*

```{r scatterplot - ggplot2 edition}

# load midwest data from the ggplot2 package
library(tidyverse)
data("midwest", package = "ggplot2")
```


Remember that each ggplot object has 3 main components:
1. The first argument in the function is the dataframe
2. The second argument takes an aes() function which defines the variables of the dataframe for each dimension of the plot

Let's try running this line of code with just the first 2 components. What do you see?

```{r}
ggplot(data = midwest, aes(x = area, y = poptotal))
```


3. The third argument tells ggplot what type of graphical geometry we want it to add
Here, we want to make a scatter plot so we'll use the argument geom_point() to visualize the data as points


```{r}
ggplot(data = midwest, aes(x = area, y = poptotal)) +
  geom_point()
```


Notice that the `+` allows us to add additional functions to the ggplot object.
The layout of the plot suggests a relationship between area and population total.
What if we wanted to reverse this graphical relationship? We could switch the variables supplied to each aesthetic or use `+ coord_flip()`

```{r}
ggplot(data = midwest, aes(x = , y = )) +
  geom_point() 
```

We still don't have all the information from our base R plot represented here, let's add some color!
Here we provide another aesthetic characteristic for just the points - their color will depend on the state.
And take a look at that nice legend that was automatically generated!

```{r}
ggplot(data = midwest, aes(x = area, y = poptotal)) +
  geom_point(aes(col = )) 
```

Now let's add more descriptive axis labels and titles
Take a look at the help page for the labs() function by typing ?labs
The code below will produce our colored plot with more descriptive x- and y-axis labels and a title


```{r}
ggplot(data = midwest, aes(x = area, y = poptotal)) +
  geom_point(aes(col = state)) +
  labs()
```


Sometimes you want to visualize several features of the data in a single plot
Let's color the points by state as before, but change the shape of the points to correspond to whether or not in a metro area
How does this change the appearance of the legend?


```{r}
ggplot(data = midwest, aes(x = area, y = poptotal)) +
  geom_point(aes(col = state, shape = )) +
  labs(title = "Midwest",
       x = "Area",
       y = "Total Population")
```


Coloring the different points by state is helpful but there are so many that it's hard to see what's going on at low populations
With facets, ggplot lets us quickly split our single plot into multiple panels, one for each state

```{r}
ggplot(data = midwest, aes(x = area, y = poptotal)) +
  geom_point() +
  facet_grid()
```

But what if we'd rather have our facets run vertically? How would we need to adjust our code?
Remember you can use `?facet_grid` to learn more about the function

```{r}
ggplot(data = midwest, aes(x = area, y = poptotal)) +
  geom_point() +
  facet_grid()
```


It seems reasonable that the relationship between size and area might differ between metropolitan areas and more rural areas
But this difference might also vary from state to state. Can we visualize all of that in one set of plots?
Create a set of plots which are faceted by both state and whether the county is in a metro area

```{r}
ggplot(data = midwest, aes(x = area, y = poptotal)) +
  geom_point() +
  facet_grid()
```

**Challenge**: How could you change the points from circles to be another shape?

Note that `fill` controls the fill color of each point and `col` controls the outline color

```{r}
ggplot(data = midwest, aes(x = area, y = poptotal)) +
  geom_point(aes(fill = state, col = state), shape = , size = ) # make points into diamonds and increase point size

```

## Plotting time series data

We can also plot continous time-series data using the `geom_line()` function
For this excercise we will use the weather data set from the `nycflights13` package, which contains time series data of weather parameters from 3 airports in NYC

*First load and take a look at the `weather` data frame to see what kind of data is available*
*Remember you will need to use the `install.packages()` function to install this package if you haven't already.*

```{r plotting time series data in ggplot2}

library(nycflights13) # load package
head(weather)
?weather # gives detail as to what the data in each column represents
```


As you can tell, this data set contains an awful lot of data, so lets subset it to focus on dates in the first half of Jan 2013

We're also going to go ahead and convert the "origins" column from a character to a factor to help with downstream plotting functions

```{r}
weather_jan <- weather %>%
  filter(day < 15 & month == 1 & year == 2013) %>%
  mutate(origin = as.factor(origin))

# Now we can create a line graph plotting total population over time
# Notice that this code contains the three arguments needed for every ggplot object
ggplot(data = weather_jan, aes(x = time_hour, y = temp)) + 
  geom_line() +
  labs(x = "Time",
       y = "Temperature")
```

How would we change the width of this line? We can use the size argument within `geom_line()` to change line width

```{r}
ggplot(data = weather_jan, aes(x = time_hour, y = temp)) + 
  geom_line() + 
  labs(x = "Time",
       y = "Temperature")
```

We may also care how temperature changes over time vary across different locations
So, we can recreate the plot to include one line for each airport by specifying the color aesthetic to represent "origin"

```{r}
ggplot(data = weather_jan, aes(x = time_hour, y = temp)) + 
  geom_line(aes(), size = 2) +
  labs(x = "Time",
       y = "Temperature")
```

How could we change the order of the airports in the legend?
One way is to reorder the levels of the corresponding factor, then re-plot. We can do this using the `fct_relevel()` function.
Use the code below to switch the order to: JFK, LGA, EWR

```{r}
weather_jan <- weather_jan %>%
  mutate(origin = fct_relevel(origin, levels = c()))

ggplot(data = weather_jan, aes(x = time_hour, y = temp)) + 
  geom_line(aes(color = origin), size = 2) +
  labs(x = "Time",
       y = "Temperature")
```


**Practice**: How would you change the title and labels of the legend? Hint: try adding a `scale_color_manual()` layer
Alternatively, you could change the labels of each factor level in the data frame when re-ordering above 

```{r}
ggplot(data = weather_jan, aes(x = time_hour, y = temp)) + 
  geom_line(aes(color = origin)) +
  labs(x = "Time",
       y = "Temperature") + # can also change title of legend according to aesthetic it corresponds to
  scale_color_manual() 
```


Let's make our temperature plot look a bit cleaner by making use of some of ggplot's preset themes

```{r}
ggplot(data = weather_jan, aes(x = time_hour, y = temp)) + 
  geom_line(aes(color = origin)) +
  labs(x = "Time",
       y = "Temperature") +
  theme_classic() # use the "classic" theme
```

Outside of the theme presets, ggplot makes use of incredibly flexible formatting arguments
The following code will change the text font and size, change the line thickness, remove the axis ticks, and remove the legend title
Try the challenges to learn how to format even more characteristics of your ggplots!

```{r}
ggplot(data = weather_jan, aes(x = time_hour, y = temp)) + 
  geom_line(aes(color = origin), size = 0.75) +
  labs(x = "Time",
       y = "Temperature") +
  theme_classic() +
  theme(text = element_text(family = "Arial", size = 20),
        axis.ticks = element_blank(),
        axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0), face = "bold", colour = "black"), 
        axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0), face = "bold", colour = "black"),
        legend.title = element_blank())
```


**Challenge**: Create a plot of humidity over time with lines colored again by airport
Hint: change the variables listed in the plot aesthetics to replace temperature data with humidity

**Challenge**: How would you change the lines to be dashed rather than solid?
Hint: use the `linetype` aesthetic and `scale_linetype_manual()` where number values correspond to different dash patterns

**Challenge**: How would you change the line opacity?
Hint: use the `alpha` asthetic - value spans from 1.0 (opaque) to 0.0 (transparent)

**Challenge**: How would you change the color and width of the axis lines?
Hint: use the `axis.line` arguments within `theme()`



If you are going to make a series of plots with the same format, you can even create your own ggplot theme object:

```{r}
my_theme <- theme_classic() + 
            theme(text = element_text(family = "Arial", size = 20),
                  axis.ticks = element_blank(),
                  axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0), face = "bold", colour = "black"), 
                  axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0), face = "bold", colour = "black"),
                  legend.title = element_blank())

ggplot(data = weather_jan, aes(x = time_hour, y = temp)) + 
  geom_line(aes(color = origin), size = 0.75) +
  labs(x = "Time",
       y = "Temperature") +
  my_theme

```

## Exploring other "geom"s

In the third part of this module, you will work in a small group to produce a new plot using the midwest dataset or any other data set we've introduced so far.

Each group will learn about a new ggplot geometry or set of functions in ggplot2 and use them to portray something interesting about the dataset.
Use the ? function as well as google to learn about your function(s).

You can start by using the ggplot cheatsheet to understand what kind of data is needed for each plot type: https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf 

Once you understand what type of data your functions allows, return to the one of the data sets and find a relationship that you want to display. 
Remember, a good place to start is with View() or str() to see what type of data are available. 

At the end of this exercise I'll ask that you send the code that produces your plot to me
(kate.sabey@uga.edu). I'll display the code and plot and each group will take ~3 minutes to explain their plot and how they created it, to the rest of the class!

```{r geom group exercise}

### Group 1 ###

# geom_col
# geom_bar


### Group 2 ###

# geom_boxplot 
# geom_violin

# Extra Challenge: can you overlay points onto the boxplot? (Hint try "+ geom_jitter()")


### Group 3 ###

# geom_area


### Group 4 ###

# geom_histogram
# geom_dotplot or geom_density


### Group 5 ###

# geom_smooth

# Extra Challenge: can you overlay the original data points behind the smooth line? (Hint: try "+ geom_point()")
# Extra Challenge: can you change the function of the smooth line? (Hint: the "method" argument)


### Group 6 ###

# scale_fill_
# scale_color_

# What can these functions change about the plot? (Hint: explore the "name", "labels", and "values" arguments)

```

## Interactive plots

The `plotly` package is an add on to ggplot2 for generating quick interactive plots. It has a few limitations (eg. the legends or other features are often poorly displayed), but the interactive feature may be useful for data exploration during an in-person meeting.

```{r interactive plots}

install.packages("plotly")
library(plotly)

# Note that ggplots can be assigned to R objects
p <- plot_ly(data = midwest, x = ~area, y = ~poptotal, type="scatter", mode="markers")

p # plot the interactive graphic

# The googleVis package also has some great plots with tooltips built in.
```

## Spatial Data

Just like other data visualizations, mapping in R can be done a few different ways. Common packages include:
* maps
* rMaps
* mapdata
* ggmap
* choroplethr
* raster
* sp
* rgdal
The last 3 (raster, sp, and rgdal) are also useful for analyzing spatial data.

But we can actually use our old friend ggplot to make some pretty nice maps.

```{r plotting maps}

# First we'll load a few extra packages to help us access and handle some more data
# Remember you will have to install these if you haven't already using the "install.packages" function
library(maps)
library(mapdata)
library(dplyr)
library(stringr)

states <- map_data("state") # We get the data for the edges of all US states from the mapdata package
counties <- map_data("county") # Then we do the same for all US counties
il_df <- subset(states, region == "illinois") # We subset the dataset of states to just the coordinates for the border of illinois
il_county <- subset(counties, region == "illinois") # We do the same for the coordinates of the borders of counties in illinois

il_pop <- subset(midwest, state = "IL") # We want to use some of that cool midwest data so we'll subset it to just the counties in IL

il_pop$subregion <-str_to_lower(il_pop$county) # Now we combine the coordinate dataset and the population size dataset together
il_county <- left_join(il_county, il_pop, by = "subregion")

# Finally we'll plot our map using with the fill of each state colored by the population of that county
# For practice, comment how each of these lines of code contribute to the final plot
ggplot(data = il_df, aes(x = long, y = lat, group = group)) +
  coord_fixed(1.3) +
  geom_polygon(data = il_county, aes(fill = poptotal), color = "white") +
  geom_polygon(color = "black", fill = NA) +
  scale_fill_gradient(trans = "log10") +
  theme_bw()

```


## Using cowplot and ggsave

Often to generate figures for publication, you will need to combine multiple plots into a single figure.
Though there are several ways to concatenate plots to generate such images, I like to use the cowplot package.

```{r merging and saving plots}

# First install and load the cowplot package if you haven't already
install.packages("cowplot")
library(cowplot)
library(here)

# Like other features in R, ggplots can be assigned to objects
# Let's assign two of our plots from above into separate objects
temp <- ggplot(data = weather_jan, aes(x = time_hour, y = temp)) + 
              geom_line(aes(color = origin), size = 0.75) +
              labs(x = "Time",
                   y = "Temperature") +
              my_theme

rain <- ggplot(data = weather_jan, aes(x = time_hour, y = precip)) + 
              geom_line(aes(color = origin), size = 0.75) +
              labs(x = "Time",
                   y = "Precipitation") +
              my_theme

# We will then use the plot_grid function in the cowplot package to combine these plot objects into a single, multi-panel figure
# One nice functionality of cowplot is that you can extract the legend shared by both plots so that it is only printed once
# Note that the rel_widths argument specifies how much of the total figure width will be dedicated to each panel

legend <- get_legend(temp)

weather_plots <- plot_grid(temp + theme(legend.position = "none"),
                           rain + theme(legend.position = "none"),
                           legend,
                           labels = c("A", "B"), label_fontfamily = "Arial", label_size = 20,
                           nrow = 1, rel_widths = c(1,1,0.3))

weather_plots

## Challenge: How would you arrange these figures vertically rather than horizontally?
## Hint: specify `ncol = 1` instead of `nrow = 1`



# We can then use the `ggsave()` function to export our figure to any location specified by a given file path

# This function also has useful arguments to manually set the file name, and desired size/resolution of the exported figure.
# Note that this function produces higher resolution ouptut than the "Export" dropdown of the "Plots" viewer panel in R studio.
# You can also use this function to save eps (Adobe illustrator) or tiff files according to specific journal specifications.

ggsave(here("weather_plots.png"), device = "png", width = 8, height = 4, units = "in", dpi = 300)

```




